Thank you, next: demographic consequences of partner diversity and turnover in a multi-species ant-plant mutualism

### 2024-08-09 Manuscript/ant_cactus_AmNat.tex
edits through end of methods

This study was conducted in the Los Pi$\tilde{n}$os mountains, a small mountain chain located on the Sevilleta National Wildlife Refuge, a Long-term Ecological Research site (SEV-LTER) in central New Mexico, USA.
This is an area characterized by steep, rocky slopes, and perennial vegetation including grasses (\textit{Bouteloua eriopoda} and \textit{B. gracilis}), yuccas, cacti, and junipers. 
Tree cholla cacti are common in high Chihuahuan desert habitats, with their native range spanning the southwestern USA \citep{Benson1982}. 
These arborescent plants produce cylindrical segments with large spines. 
In the growing season (May to August in New Mexico), the plants initiate new vegetative segments and flower buds at the ends of existing segments. 
While most plants produce new segments every season, only those which are reproductively mature produce flower buds. 
%Tree cholla generally reach at least 9 years of age before beginning to reproduce \citep{Ohm2014}.
Like other EFN-bearing cacti, tree cholla secrete nectar from specialized glands on young vegetative segments and flower buds \citep{Ness2006,Oliveira1999}. 
Flower buds produce more and higher-quality EFN than vegetative segments, making reproductive cholla valuable mutualist partners \citep{Miller2014}. 
%Smaller cholla produce little to no EFN, so larger cholla, especially flowering individuals, are generally more highly tended.

Tree cholla EFN is harvested by various ant species. 
At SEV-LTER, cholla are visited primarily by two species of ground-nesting ants, \textit{Crematogaster opuntiae} and \textit{Liometopum apiculatum}, as well as other rarer species, including \textit{Forelius pruinosus} and unidentified species in the genera \textit{Aphaenogaster} and \textit{Camponotus}.
\textit{L. apiculatum} are the most frequent visitors with $25\% - 60\%$ of tree cholla tended by these ants, followed by \textit{C. opuntiae} visiting between $5\% - 20\%$ of cacti depending on the year \citep{Donald2022}. 
Between $ 30\% - 80\%$ of cacti remain vacant in any given year. 
Workers of different species rarely co-occur on individual plants, likely due to interspecific competition \citep{Miller2007}: staged introductions of \textit{C. opuntiae} to \textit{L. apiculatum}-tended plants, and vice versa, provoke aggressive responses by residents (A. Cambpell, \textit{personal observation}).
%Each cholla is visited by a single ant species for the duration of a season, and the species of the visitors can change from one season to the next. 
In Fall, tree cholla stop producing EFN and the ants vacate until the next growing season.

Multiple insect herbivores and seed predators specialize on tree cholla \citep{Mann1969}. 
The Cerambycid beetle \textit{Moneilema appressum} and a weevil (Coleoptera: Curculionidae) of the genus \textit{Gerstaekeria} feed on vegetative and reproductive structures as adults and their larvae feed internally. 
A cactus bug, \textit{Narnia pallidicornis} (Hemiptera: Coreidae), feeds on all cholla parts with a preference for the reproductive structures \citep{Miller2006}.
A seed predator, \textit{Cahela ponderosella} (Lepidoptera: Pyralidae), oviposits in open flowers and larvae eat seeds in developing fruits. 
These predators can have significant negative impacts on plant fitness of and depress population growth \citep{Miller2009}.
Prior experiments showed that ant-tended tree cholla experience less herbivory and seed predation than plants from which ants were excluded \citep{Miller2007,Ohm2014}.

\subsection*{Data Collection}
This study is based on long-term demographic data spanning 2004 to 2023 at SEV-LTER. 
From 2004 to 2008, we censused 134 plants distributed across three spatial blocks. 
This initial census group was discontinued in 2009, when we established six 30 $\times$ 30-meter plots and tagged all tree cholla within those plots. 
Two additional 30 $\times$ 30-meter plots were added in 2011, and this group of eight plots has since been censused annually through 2023 (with the exception of 2020 due to the pandemic shutdown). 
For all plants, in May or early June of each year we recorded plant survival since the last survey and, for survivors, we recorded height (cm), maximum crown width (cm), and crown width perpendicular to the maximum (cm).
Size measurements were used to calculate plant volume ($cm^3$) based on the volume of an elliptical cone. 
We measured reproduction as counts of viable and aborted flowerbuds. 
We recorded the ant species present (or vacancy if no ants present).
Occurrences of more than one ant species on one plant were rare (less than 5\% of observations), and for the purpose of this analysis we classified the plant as being occupied by the more abundant species. 
Plots were searched for new recruits each year, and these were added to the census.
In total, the data set included 1141 unique individuals and 19 year observations. 
%These data were used to fit vital rate models (survival, growth, reproduction) as functions of plant size and ant occupancy state.

@@

\subsection*{Multi-state Integral Projection Model}
Integral Projection Models describe population dynamics in discrete time, with functions that relate vital rates to continuous state variables \citep{ellnerbook}. 
While IPMs are a natural choice for populations with continuous size structure, they can also be modified to accommodate a combination of continuous and discrete state variables, as we do here. 
We constructed a stochastic, multi-state IPM that stitches together population structure associated with plant size and ant state, allowing us to determine the individual fitness effects of each ant species and the composite effects of multiple partners, with ant transition dynamics and inter-annual variability modeled explictly.

@@

Ant state is included as a predictor variable in IPM sub-models where there are biologically realistic pathways through which ants could impact the outcome of that process. 
For example, ant partners defend cacti from herbivores, and prior experimental work indicates that ant tending can reduce vegetative tissue loss and floral abortion.
Therefore, ant state was included in sub-models for survival, growth, and flowerbud viability. 
In contrast, we have no reason to expect that ant tending can directly influence the probability of flowering or flowerbud production independently of its influence on plant size, so these sub-models do not include ant state as a predictor variable.

We modeled the tree cholla life cycle using continuously size-structured plants where $n(x,a)_{t}$ gives the number of plants of size $x$  and ant state $a$ in year $t$, plus two discrete seed banks ($B^1_{t}$ and $B^2_{t}$) corresponding to 1 and 2-year old seeds. 
Seed bank dynamics are given by:

@@

\noindent %In these equations $x$ and $x'$ indicate the size of a plant in years $t$ and $t+1$ respectively, $a$ and $a'$ indicate the ant partner of a plant in years $t$ and $t+1$.
Functions $P(x;\pmb{\tau}^P)$ and $F(x;\pmb{\tau}^F)$ give the probability of flowering in year $t$ and the number of flowerbuds produced in year $t$, respectively, by plants of size $x$. 
The proportion of flowerbuds that remain viable through fruit set ($V(a;\pmb{\tau}^V_{a})$) and the number of seeds per fruit ($\kappa(a')$) is dependent on ant state $a$. 
The vectors $\pmb{\tau}$ give year-specific deviates (mean zero) and appear in functions for which we can estimate temporal stochasticity from the long-term data; superscripts indicate the corresponding vital rate and, when present, the $a$ subscript indicates that deviates are specific to plants in ant state $a$.
For example, temporal deviates $\pmb{\tau}^V_{a}$ describe better- and worse-than-average years for flowerbud viability and plants in different ant states can fluctuate independently (good years for \textit{L. apiculatum} -occupied plants may not be good years for \textit{C. opuntiae}-occupied plants, for example). 
Seed production is integrated over the size distribution, from the lower $L$ to upper $U$ size limits, and summed over all possible ant states ($A=4$) giving total seed production. 
Seeds are multiplied by the probability of seed dispersal and survival ($\delta$) to give the number of seeds that enter the one-year old seed bank. 
Plants can recruit out of the year-one seed bank with probability $\gamma_1$ or transition to the two-year seed bank with a probability of $1 - \gamma_1$. 
Seeds in the two-year seed bank are assumed to either germinate with probability $\gamma_2$ or die.

For the above-ground part of the life cycle, the number of plants of size $x'$ and ant state $a'$ in year $t+1$ ($n(x',a')_{t+1}$) is given by survival/growth transitions from size $x$ and ant state $a$ in year $t$, plus germination out of the seed banks:
\begin{linenomath*}
	$$
	n(x',a')_{t+1} = (\gamma_1 B^1_{t} + \gamma_2 B^2_{t} ) \eta(x') \omega \rho_{0}(a')  + \\
	$$
	$$
	\sum_{a=1}^{A} \int_L^U S(x,a;\pmb{\tau^S_{a}}) G(x',x,a;\pmb{\tau^G_{a}}) \rho(x,a,a';\pmb{\tau^{\epsilon}}) n(x,a)_t dx \\
	$$
	\label{eqn:IPM2}
\end{linenomath*}

\noindent The first term in Eq. \tom{\ref{eqn:IPM2}}{We should label equations. I am not sure why the equation label is not working here and I did not try to figure it out. It is probably something with the linenomath formatting.} estimates the number of individuals recruiting from a one or two-year seed bank to a plant of size $x'$ and ant state $a'$ based on the recruit size distribution $\eta(x')$ and the probability of seedling survival ($\omega$) from germination (late summer) to the census (May).
This term is multiplied by $\rho_{0}(a')$, which gives the probability that a new recruit has ant state $a'$ at its first appearance in our census ($\sum\rho_{0}(a')=1$). 
The second term represents all possible transitions from size $x$ and ant $a$ to size $x'$ and ant $a'$, conditioned on survival. 
Survival ($S(x,a;\pmb{\tau}^S_{a})$) and growth from size $x$ to $x'$ ($G(x',x,a;\pmb{\tau}^G_{a})$) are both dependent on initial size and ant state. 
As above, these functions include inter-annual variability through year-specific deviates that can vary by ant state ($\pmb{\tau}_{a}$). 
Finally, ant transition function $\rho(a',a,x;\pmb{\tau}^{\rho})$ gives the probability that an individual transitions from ant state $a$ to $a'$ in the next census, conditional on initial size $x$. 
This function includes inter-annual variability through year-specific intercepts which are consistent across ant states ($\pmb{\tau}^\rho$).

\subsection*{Statistical modeling and parameter estimation}
We parameterized the IPM using a series of generalized linear mixed models in a hierarchical Bayesian framework. 
Vital rate models included spatial and temporal random effects associated with plot and year variation, respectively (only year variation is used in the IPM), and included plant size ($log(cm^3)$; $x,x'$), ant partner state ($a,a'$), or both as fixed-effect predictor variables. 
In addition to vital rate models describing plant demographic performance, we also fit a sub-model to predict transitions between ant states conditional on plant size and previous ant state. 
As in the IPM, our statistical modeling assumed that demographic effects of ant occupancy are limited to survival, growth, and flowerbud viability.

@@

\paragraph{Growth}
We fit the growth sub-model ($G(x',x,a;\pmb{\tau^G_{a}})$) to data on size in year $t+1$ ($y^G$) using the skewed normal distribution to account left-skewed size transitions:
$$y_i^G \sim Skewed Normal(\hat{G_i},\sigma_i,\alpha_i) $$
$$\hat{G_i} = \beta^0_{a[i]} + \beta^1_{a[i]} x_i + \beta^2_{a[i]} x_i^2 + u_{year[i],a[i]} + w_{plot[i]} $$
The mean of the $i$th observation $\hat{G_i}$ is a second-order polynomial with ant-size interactions because  preliminary analysis found this was an improvement over a linear relationship. 
The year- and ant-specific random effect $u$ (which parameterizes the $\pmb{\tau}^G_{a}$ vector) and plot-specific random effect $w$ are normally distributed with variances $\sigma^2_{year}$ and $\sigma^2_{plot}$, respectively. 
Parameters $\sigma_i$ and $\alpha_i$  control residual variance and skewness, respectively, and were defined as linear functions of initial size $x_i$ ($\sigma_i$ is strictly positive and was modeled with a log link function). 
We assume growth variance and skewness were not dependent on ant occupancy state.

\paragraph{Survival}
The survival sub-model ($S(a,x;\pmb{\tau}_{a}^{S})$) estimates the probability of survival from year $t$ to year $t+1$, with fixed effects of size $x$ and ant partner $a$ in year $t$.
We fit this model to the survival data (alive or dead) using a Bernoulli distribution and the logit link function, with a similar linear predictor as the growth model but without the second-order influence of size.

\paragraph{Reproduction}
The flowering sub-model ($P(x;\pmb{\tau}^{P})$) estimates the probability of reproducing in year $t$, with fixed effects for the size $x$ and random effects of plot and year.
We fit this model to the reproductive status data (vegetative or flowering) using a Bernoulli distribution and a logit link function, similar to the survival model above but with no ant effects.  
The flower bud function $F(x;\pmb{\tau}^{F})$ estimates the total flowers produced by a reproducing plant in year $t$, with fixed effects of size $x$. 
We fit this model to flowerbud count data (sum of viable and aborted buds) using a zero-truncated negative binomial distribution with a log link and normally distributed year and plot random effects.

The flowerbud viability sub-model ($V(a;\pmb{\tau}^{V}_{a})$) estimates the proportion of flowers produced by a plant that are viable (not aborted) in year $t$, with fixed effects of ant partner $a$ in year $t$.
We fit this model to floral viability data using a binomial distribution where trials and successes are given by the total number of flower buds and the number that are viable, respectively.
This model used a logit link function and included random effects for plot and year following the same structure as the growth and survival models.

Additional reproductive parameters for the number of seeds per fruit, probability of entry to the seed bank, germination rates, and recruit size were estimated following methods described in Appendix XX.
Estimates for the number of seeds per fruit were ant specific, but the data (cite tom's data) only included \textit{L. apiculatum} and \textit{C. opuntiae} ants.
We used the vacancy estimates for plants tended by other ants. 
The rest of these parameters were not ant specific.

\paragraph{Ant Transitions}
The ant transition model ($\epsilon(x,a,a';\pmb{\tau}^{\epsilon})$) estimates the probability of a cactus being occupied by ant partner $a'$ in year $t+1$, with fixed effects of the previous size $x$  and the previous ant partner $a$  in year $t$.
We fit this model to ant partner data using a multinomial distribution with a logit link function and random effects for plot and year.

We used this model to estimate probabilities which are counterfactual (such as the probability of being \textit{L. apiculatum} tended or vacant with no other options). 
To do this we assumed that the level of vacancy was constant across all states and that whatever ants were present in our simulation would tend the gaps which would otherwise be tended by another ant species. 
This assumes that competitive exclusion is the mechanism which determines the ants we see tending cacti, other potential approaches are outlined in Apendix XX.

\paragraph{Parameter estimation}
We fit models using Markov chain Monte Carlo (MCMC) simulations via STAN run through version 4.0.2 of R \cite{Rcite,Rstancite}. 
We used vague priors for all parameters. 
For each model, we obtained three chains of 10,000 iterations, discarding the first 1,500 iterations. 
%We did not thin the chains, thus all samples were retained following burn-in. 
We visually assessed parameter convergence between and within chains (Figures \ref{fig:Surv_post} -- \ref{fig:Pre_Surv_post} b) and assessed overall model fit with posterior predictive checks to examine how well the fitted model can generate simulated data similar to the real data (Figures \ref{fig:Surv_post} -- \ref{fig:Pre_Surv_post} a).
%All estimated parameters are available in Table \ref{tab:Params}.

\subsection*{IPM Analysis}
Analyzing an IPM requires discretizing the continuous IPM kernel into an approximating matrix. 
Size variable $x$ is discretized into $b$ bins, resulting in a $b \times b$ matrix.
In our model there is additional complexity in the form of transitions between $A$ ant partners and two additional discrete states (year one and year two seed banks), leading to a matrix size of $A(b+2) \times A(b+2)$.
We used $b = 200$ bins (sufficient for numerically stable model outputs) and extended the integration limits beyond the minimum ($L$) and maximum ($U$) observed sizes to avoid unintentional eviction \cite{Williams2012}.

For stochastic analyses, we estimated the approximating matrix corresponding to each $t$ to $t+1$ transition year. 
To estimate population mean fitness in a stochastic environment ($\lambda_{S}$) we simulated population dynamics for 500 years by randomly sampling among the 17 annual transition matrices, discarding the first 100 years of the simulation to minimize the influence of initial conditions. 
Sampling observed transition matrices (rather than independently sampling regression coefficients) produces models that realistically capture inter-annual variation by preserving correlations between vital rates \cite{metcalf2015statistical}.
We tallied the total population size at each time step as  $N_{t} = B^1_{t} + B^2_{t} + \sum_{a=1}^{A}\int n(x,a)_{t}dx$ and calculated the stochastic growth rate as $log(\lambda_S) = E[log(\frac{N_{t}}{N_{t+1}})]$ \citep{Mark 2009}.
We propagated uncertainty from the vital rate models using 1000 draws from the joint posterior distribution of model parameters, resulting in a posterior distribution of $\lambda_{S}$ and other derived quantities.

\subsubsection*{Partner diversity experiments}
Using the fully parameterized multi-state IPM, we conducted simulation ``experiments'' to quantify how diversity and identity of ant partners influenced plant fitness ($\lambda_{S}$). 
From the fullest version of the model corresponding to the observed assemblage of partners, we created subsets corresponding to all eight possible scenarios of diversity and composition: no ant partners (complete vacancy); one ant partner (\textit{C. opuntiae} only, \textit{L. apiculatum} only, Other only); two partners (all pairwise combinations of \textit{C. opuntiae}, \textit{L. apiculatum}, and Other); and three partners (observed scenario of all ant states).
These simulation experiments were made possible by extrapolating ant-specific demographic performance across the size distribution, even for combinations of size and ant occupancy that were rarely or never observed. 
For example, the no-partner scenario modeled a hypothetically ant-free cactus population, even though no such population exists to our knowledge, by applying the statistical knowledge gleaned from vacant plants (which were mostly small and non-reproductive) across the size distribution.

>>>>

This study was conducted in the Los Pi$\tilde{n}$os mountains, a small mountain chain located on the Sevilleta National Wildlife Refuge, a Long-term Ecological Research site (SEV-LTER) in central New Mexico, USA.
This is an area characterized by steep, rocky slopes, and perennial vegetation including grasses (\textit{Bouteloua eriopoda} and \textit{B. gracilis}), yuccas, cacti, and junipers. 
Tree cholla cacti are common in high Chihuahuan desert habitats of the southwestern USA \citep{Benson1982}. 
These arborescent plants produce cylindrical segments with large spines. 
In the growing season (May to August in New Mexico), the plants initiate new vegetative segments and flower buds at the ends of existing segments. 
While most plants produce new segments every season, only those which are reproductively mature produce flower buds. 
%Tree cholla generally reach at least 9 years of age before beginning to reproduce \citep{Ohm2014}.
Like other EFN-bearing cacti, tree cholla secrete nectar from specialized glands on young vegetative segments and flower buds \citep{Ness2006,Oliveira1999}. 
Flower buds produce more and higher-quality EFN than vegetative segments, making reproductive cholla valuable mutualist partners \citep{Miller2014}. 
Smaller, non-flowering cholla produce little to no EFN and are commonly vacant (no ant visitation at the time of our census) \citep{Miller2014}.

Tree cholla EFN is harvested by various ground-nesting ant species. 
At SEV-LTER, cholla are visited primarily by two species, \textit{Crematogaster opuntiae} and \textit{Liometopum apiculatum}, as well as other rarer species, including \textit{Forelius pruinosus} and unidentified species in the genera \textit{Aphaenogaster} and \textit{Camponotus}.
\textit{L. apiculatum} are the most frequent visitors with $25\% - 60\%$ of tree cholla tended by these ants in a given year, followed by \textit{C. opuntiae} ($5\% - 20\%$) \citep{Donald2022}. 
Between $ 30\% - 80\%$ of cacti are vacant in any given year. 
Workers of different species rarely co-occur on individual plants, likely due to interspecific competition. 
For example, staged introductions of \textit{C. opuntiae} to \textit{L. apiculatum}-tended plants, and vice versa, provoke aggressive responses by residents (A. Cambpell, \textit{personal observation}).
%Each cholla is visited by a single ant species for the duration of a season, and the species of the visitors can change from one season to the next. 
%In Fall, tree cholla stop producing EFN and the ants vacate until the next growing season.

Several insect herbivores and seed predators specialize on tree cholla \citep{Mann1969}, and defense against these enemies is the main pathway by which ant visitation affects plant demography. 
The Cerambycid beetle \textit{Moneilema appressum} and an unidentified weevil (Coleoptera: Curculionidae) of the genus \textit{Gerstaekeria} feed on vegetative and reproductive structures as adults and their larvae feed internally. 
Two species of cactus bugs, \textit{Narnia pallidicornis} and \textit{Chelinidea vittiger} (Hemiptera: Coreidae), feed on all cholla parts with a preference for flower buds; their damage can induce floral abortion \citep{Miller2006}. 
A seed predator, \textit{Cahela ponderosella} (Lepidoptera: Pyralidae), oviposits in open flowers and larvae eat seeds in developing fruits. 
These consumers can have significant negative impacts on plant fitness and depress population growth \citep{Miller2009}.
Prior experiments showed that ant-tended tree cholla experience less herbivory and seed predation than plants from which ants were excluded \citep{Miller2007,Ohm2014}.

\subsection*{Data Collection}
This study is based on long-term demographic data spanning 2004 to 2023 at SEV-LTER. 
From 2004 to 2008, we censused 134 plants distributed across three spatial blocks. 
This initial census group was discontinued in 2009, when we established six 30 $\times$ 30-meter plots and tagged all tree cholla within those plots. 
Two additional 30 $\times$ 30-meter plots were added in 2011, and this group of eight plots has since been censused annually through 2023 (with the exception of 2020 due to the pandemic shutdown). 
For all plants, in May or early June of each year we recorded plant survival since the last survey and, for survivors, we recorded height (cm), maximum crown width (cm), and crown width perpendicular to the maximum (cm).
Size measurements were used to calculate plant volume ($cm^3$) based on the volume of an elliptical cone. 
We measured reproduction by counting flowerbuds, and we distinguished between flowerbuds that were viable and aborted. 
We recorded the ant species present (or vacancy if no ants present).
Occurrences of more than one ant species on one plant were rare (less than 5\% of observations), and for the purpose of this analysis we classified the plant as being occupied by the more abundant species. 
Plots were searched for new recruits each year, and these were added to the census.
These data allowed us to link each plant's demographic fate (survival, growth, and reproduction) to its state of ant visitation. 
In total, the data set includes a total of 9,787 observations of 1141 unique individuals across \tom{19}{There are actually 16 years (you can see my tally in the 01 script). In your code you took length(unique(Yeart1)). But one of these years is 2024 (for which there is no t1 data entered), one is NA, and 2019-2021 does not count.} complete transition years (spanning May/June of year $t$ to May/June of year $t+1$). 
\tom{}{You should cite the data package here. The published package only goes up to 2018, so we should talk about updating that package with you as lead author, which should be easy to do as long as the data are QAQC'd.}
%These data were used to fit vital rate models (survival, growth, reproduction) as functions of plant size and ant occupancy state.

@@

\subsection*{Multi-state Integral Projection Model}
The demographic data were used to parameterize a multi-state Integral Projection Model (IPM).
IPMs describe population dynamics in discrete time with functions that relate vital rates to continuous state variables, typically size \citep{ellnerbook}. 
While IPMs are a natural choice for populations with continuous size structure, they can also be modified to accommodate a combination of continuous and discrete state variables, as we do here. 
We constructed a stochastic, multi-state IPM that stitches together population structure associated with plant size and ant state, allowing us to determine the individual fitness effects of each ant species and the composite effects of multiple partners, with ant transition dynamics and inter-annual variability modeled explictly.

@@

Ant state is included as a predictor variable in IPM sub-models where there are biologically realistic pathways through which ants could impact the outcome of that process. 
For example, ant partners defend cacti from herbivores, and prior experimental work indicates that ant tending can reduce vegetative tissue loss and floral abortion.
Therefore, ant state was included in sub-models for survival, growth, flowerbud viability, and seed number per flowerbud. 
In contrast, we have no reason to expect that ant tending can directly influence the probability of flowering or flowerbud production independently of its influence on plant size, so these sub-models do not include ant state as a predictor variable.

We modeled the tree cholla life cycle using continuously size-structured plants where number of plants of size $x$  and ant state $a$ in year $t$ ($n(x,a)_{t}$) predicts the number of plants of size $x'$ and ant state $a'$ in year $t+1$ ($n(x',a')_{t+1}$) based on a size- and ant-specific vital rates. 
The models also includes two discrete seed banks ($B^1_{t}$ and $B^2_{t}$) corresponding to 1 and 2-year old seeds. 
Seed bank dynamics are given by:

@@

\noindent %In these equations $x$ and $x'$ indicate the size of a plant in years $t$ and $t+1$ respectively, $a$ and $a'$ indicate the ant partner of a plant in years $t$ and $t+1$.
Functions $P(x;\pmb{\tau}^P)$ and $F(x;\pmb{\tau}^F)$ give the probability of flowering in year $t$ and the number of flowerbuds produced in year $t$, respectively, by plants of size $x$ in year $t$. 
The proportion of flowerbuds that remain viable through fruit set ($V(a;\pmb{\tau}^V_{a})$) and the number of seeds per fruit ($\kappa(a')$) is \tom{dependent on ant state $a$}{The text says ant state $a$ but the notation here and in the equation shows ant state $a'$. Which is it? (Should be $a$.)}. 
The vectors $\pmb{\tau}$ give year-specific deviates (mean zero) and appear in functions for which we can estimate temporal stochasticity from the long-term data; superscripts indicate the corresponding vital rate and, when present, the $a$ subscript indicates that deviates are specific to plants in ant state $a$.
For example, temporal deviates $\pmb{\tau}^V_{a}$ describe better- and worse-than-average years for flowerbud viability and plants in different ant states can fluctuate independently (good years for \textit{L. apiculatum} -occupied plants may not be good years for \textit{C. opuntiae}-occupied plants, for example). 
Seed production is integrated over the size distribution, from the lower $L$ to upper $U$ size limits, and summed over all possible initial ant states ($A=4$) giving total seed production. 
Seeds are multiplied by the probability of escaping post-dispersal seed predation ($\delta$) to give the number of seeds that enter the one-year old seed bank. 
Plants can recruit out of the year-one seed bank with probability $\gamma_1$ or transition to the two-year seed bank with a probability of $1 - \gamma_1$. 
Seeds in the two-year seed bank are assumed to either germinate with probability $\gamma_2$ or die.

For the above-ground part of the life cycle, the number of plants of size $x'$ and ant state $a'$ in year $t+1$ is given by survival/growth transitions from size $x$ and ant state $a$ in year $t$, plus germination out of the seed banks:
\begin{linenomath*}
	$$
	n(x',a')_{t+1} = (\gamma_1 B^1_{t} + \gamma_2 B^2_{t} ) \eta(x') \omega \rho_{0}(a')  + \\
	$$
	$$
	\sum_{a=1}^{A} \int_L^U S(x,a;\pmb{\tau^S_{a}}) G(x',x,a;\pmb{\tau^G_{a}}) \rho(x,a,a';\pmb{\tau^{\epsilon}}) n(x,a)_t dx \\
	$$
	\label{eqn:IPM2}
\end{linenomath*}

\noindent The first term in Eq. \tom{\ref{eqn:IPM2}}{We should label equations. I am not sure why the equation label is not working here and I did not try to figure it out. It is probably something with the linenomath formatting.} estimates the number of individuals recruiting from a one or two-year seed bank to a plant of size $x'$ and ant state $a'$ based on the recruit size distribution $\eta(x')$ and the probability of over-winter seedling survival ($\omega$) from germination (late summer) to the census (May).
This term is multiplied by $\rho_{0}(a')$, which gives the probability that a new recruit has ant state $a'$ ($\sum\rho_{0}(a')=1$). 
The second term represents all possible transitions from size $x$ and ant $a$ to size $x'$ and ant $a'$, conditioned on survival. 
Survival ($S(x,a;\pmb{\tau}^S_{a})$) and growth from size $x$ to $x'$ ($G(x',x,a;\pmb{\tau}^G_{a})$) are both dependent on initial size and ant state. 
As above, these functions include inter-annual variability through year-specific deviates that can vary by ant state ($\pmb{\tau}_{a}$). 
Finally, ant transition function $\rho(a',a,x;\pmb{\tau}^{\rho})$ gives the probability that an individual transitions from ant state $a$ to $a'$ in the next census, conditional on initial size $x$. 
This function includes inter-annual variability through year-specific intercepts which are consistent across initial ant states ($\pmb{\tau}^\rho$).

\subsection*{Statistical modeling and parameter estimation}
We parameterized the IPM using a series of generalized linear mixed models in a hierarchical Bayesian framework. 
Vital rate models included spatial and temporal random effects associated with plot and year variation, respectively (only year variation is used in the IPM), and included plant size (the natural logarithm of volume, $log(cm^3)$; $x,x'$), ant partner state ($a,a'$), or both as fixed-effect predictor variables. 
%In addition to vital rate models describing plant demographic performance, we also fit a sub-model to predict transitions between ant states conditional on plant size and previous ant state. 
As in the IPM, our statistical modeling assumed that demographic effects of ant occupancy are limited to survival, growth, and flowerbud viability.

@@

\paragraph{Growth}
We fit the growth sub-model ($G(x',x,a;\pmb{\tau^G_{a}})$) to data on size in year $t+1$ ($y^G$) using the skewed normal distribution to account for left-skewed size transitions (at some initial sizes, transitions below the expected future size were more common than transitions above it). 
The skew-normal has three parameters corresponding to location ($\hat{G}$), shape ($\sigma$), and skew ($\alpha$):
$$y_i^G \sim Skewed Normal(\hat{G_i},\sigma_i,\alpha_i) $$
$$\hat{G_i} = \beta^0_{a[i]} + \beta^1_{a[i]} x_i + \beta^2_{a[i]} x_i^2 + u_{year[i],a[i]} + w_{plot[i]} $$
Here, the location parameter for the $i$th observation $\hat{G_i}$ is defined as a second-order polynomial with ant-size interactions because  preliminary analysis found this was an improvement over a linear relationship (note that the location parameter of the skew-normal is not the mean). 
The year- and ant-specific random effect $u$ (which parameterizes the $\pmb{\tau}^G_{a}$ vectors) and plot-specific random effect $w$ are normally distributed with variances $\sigma^2_{year}$ and $\sigma^2_{plot}$, respectively. 
Parameters $\sigma_i$ and $\alpha_i$  control residual variance and skewness, respectively, and were defined as linear functions of initial size $x_i$ ($\sigma_i$ is strictly positive and was modeled with a log link function). 
We assume growth variance and skewness were not dependent on ant occupancy state.

\paragraph{Survival}
The survival sub-model ($S(a,x;\pmb{\tau}_{a}^{S})$) estimates the probability of survival from year $t$ to year $t+1$, with fixed effects of size $x$ and ant partner $a$ in year $t$.
We fit this model to the survival data (alive or dead) using a Bernoulli distribution with a similar linear predictor for the probability of survival as in the growth model but with a logit link function and without the second-order influence of size.

\paragraph{Reproduction}
The flowering sub-model ($P(x;\pmb{\tau}^{P})$) estimates the probability of reproducing in year $t$, with fixed effects size $x$ in year $t$ and random effects of plot and year.
We fit this model to the reproductive status data (vegetative or flowering) using a Bernoulli distribution and a logit link function, similar to the survival model above but with no ant effects.  
The flower bud function $F(x;\pmb{\tau}^{F})$ estimates the total flowers produced by a reproducing plant in year $t$, with fixed effects of size $x$ in year $t$. 
We fit this model to flowerbud count data (sum of viable and aborted buds) using a zero-truncated negative binomial distribution with a log link and normally distributed year and plot random effects.

The flowerbud viability sub-model ($V(a;\pmb{\tau}^{V}_{a})$) estimates the probability that flowerbuds produced in year $t$ remain viable (not aborted), with fixed effects of ant partner $a$ in year $t$.
We fit this model to floral viability data using a binomial distribution where trials and successes are given by the total number of \tom{flower buds}{Sometimes this is flowerbuds and sometimes it is flower buds. I would tighten this up.} and the number that are viable, respectively.
This model used a logit link function and included random effects for plot and year following the same structure as the growth and survival models, with ant-specific year random effects.

Additional reproductive parameters for the number of seeds per fruit, probability of entry to the seed bank, germination rates, and recruit size were estimated following methods described in Appendix XX.
Estimates for the number of seeds per fruit were ant specific, but the data (cite tom's data) only included \textit{L. apiculatum} and \textit{C. opuntiae} ants.
\tom{We used the vacancy estimates for plants tended by other ants. }{This is an assumption that needs to be explained or justified. Reviewers will ask if results are sensitive to this assumption.}
\tom{The rest of these parameters were not ant specific. }{Because the seed number model is ant-specific, I suggest moving it from the appendix to the main text.}

\paragraph{Ant Transitions}
The ant transition model ($\epsilon(x,a,a';\pmb{\tau}^{\epsilon})$) estimates the probability of a cactus being occupied by ant partner $a'$ in year $t+1$ given that it was occupied by initial ant partner $a$  in year $t$, with fixed effect of initial size $x$.
We fit this model to ant partner data using a multinomial distribution with a logit link function and random effects for plot and year.

\paragraph{Parameter estimation}
We fit models using Stan run through version 4.0.2 of R \cite{Rcite,Rstancite}. 
We used vague priors for all parameters. 
For each model, we obtained three chains of 10,000 iterations, discarding the first 1,500 iterations. 
%We did not thin the chains, thus all samples were retained following burn-in. 
\tom{We visually assessed parameter convergence between and within chains (Figures \ref{fig:Surv_post} -- \ref{fig:Pre_Surv_post} b) and assessed overall model fit with posterior predictive checks to examine how well the fitted model can generate simulated data similar to the real data (Figures \ref{fig:Surv_post} -- \ref{fig:Pre_Surv_post} a).}{These figures need to be labeled so that it is clear that they are in the Appendix. Am Nat's template should allow you to do this automatically.}
%All estimated parameters are available in Table \ref{tab:Params}.

\subsection*{IPM Analysis}
Analyzing an IPM requires that we discretize the continuous IPM kernel into an approximating matrix. 
Size variable $x$ was discretized into $b$ bins, resulting in a $b \times b$ matrix.
In our model there is additional complexity in the form of transitions between $A$ ant states and two additional discrete states (year one and year two seed banks), leading to a matrix size of $A(b+2) \times A(b+2)$.
We used $b = 200$ bins, which we found to be sufficient for numerically stable outputs, and extended the integration limits beyond the minimum ($L$) and maximum ($U$) observed sizes to avoid unintentional eviction using the ``floor-and-ceiling'' method \cite{Williams2012}.

For stochastic analyses, we estimated the approximating matrix corresponding to each $t$ to $t+1$ transition year. 
To estimate population mean fitness in a stochastic environment ($\lambda_{S}$) we simulated population dynamics for 500 years by randomly sampling among the \tom{17 annual transition matrices}{Unless you add 2024 I think this is 16.}, discarding the first 100 years of the simulation to minimize the influence of initial conditions. 
Sampling observed transition matrices (rather than independently sampling regression coefficients) produces demographic time series that realistically capture inter-annual variation by preserving correlations between vital rates \cite{metcalf2015statistical}.
We tallied the total population size at each time step as  $N_{t} = B^1_{t} + B^2_{t} + \sum_{a=1}^{A}\int n(x,a)_{t}dx$ and calculated the stochastic growth rate as $log(\lambda_S) = E[log(\frac{N_{t+1}}{N_{t}})]$ \citep{Mark2009}.
We propagated uncertainty from the vital rate models using \tom{1000 draws}{I did not realize you were doing this many. I think you could use as few as 100, which would make your numerical work much faster.} from the joint posterior distribution of model parameters, resulting in a posterior distribution of $\lambda_{S}$ and other derived quantities.
